Covering Your Basis: Model-Agnostic Inference of LIGO-Virgos Binary Black Hole Spin Distribution

### 2022-09-14 src/tex/intro.tex
update the git repo

\documentclass[twocolumn, linenumber]{aastex63}

\newcommand{\msun}{\ensuremath{{\rm M}_\odot}}
\newcommand{\result}[1]{\textcolor{red}{#1}}
\newcommand{\bruce}[1]{\textcolor{blue}{BE: #1}}
\newcommand{\NewChange}[1]{\textcolor{green}{#1}}

\graphicspath{{./}{figures/}}

\usepackage{showyourwork}
\usepackage[perpage]{footmisc}
\usepackage{colortbl}
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{appendix}
\definecolor{Gray}{gray}{0.9}
\linenumbers

\begin{document}

\title{Covering Your Basis: Model-Agnostic Inference of LIGO-Virgos Binary Black Hole Spin Distribution}

\author{Bruce Edelman}
\email{bedelman@uoregon.edu}
\affiliation{Institute  for  Fundamental  Science, Department of Physics, University of Oregon, Eugene, OR 97403, USA}

\author{Ben Farr}
\affiliation{Institute  for  Fundamental  Science, Department of Physics, University of Oregon, Eugene, OR 97403, USA}

\begin{abstract}
We introduce a non-parametric model built from basis splines, and apply it to infer the astrophysical mass distribution of 
binary black holes using GWTC-3.
\end{abstract}

@@

Now that gravitational wave observations of compact binary coalescence is a regular occurrence, producing catalogs of nearly 
100 such detections, we are beginning to have access to a more complete understanding of the population of compact binaries. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA collaboration, the catalog size is expected to increase by up to a factor of 5. 
Looking at population properties of compact objects lets us probe the complex physics common to all of our observations that governs the 
formation and evolution of merging compact objects. The history of these dense objects is imprinted on the properties of the population 
of CBCs, allowing us to learn about these not well understood ideas through careful evaluation of the enture distribution of observed GWs. 
The mass distribution of binary black holes has been used to measure cosmological parameters, constrain modified gravitational wave propagation, 
constrain a non-constant Planck mass, look for evidence of ultralight bosons through superradiance, put constraints on the uncertain nuclear reaction 
rates important for stellar evolution, look for primordial black holes, to constrain the neutron star equation of state, 
and to understand the fromation channels/environments of compact binary coalescences (CBCs). Through a better understanding of the mass, spin, 
and redshift distribution of compact binaries that will come with the increased catalog size, one can probe these dufferent physical phenoma with 
greater fidelity.

@@

\section{Methods} \label{sec:methods}
\subsection{Constructing A Basis} \label{sec:basis_splines}

A common non-parametric method in statistical applications is the use of B-Splines or basis splines. A spline function of order $k$, 
is a piece-wise polynomial of order $k-1$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
B-Splines of order $k$ are a set of order $k-1$ polynomials that form a complete basis for any spline function of order $k$. 
Given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of B-Splines for 
every possible spline function with knots, $\mathbf{t}$. One can construct the B-Splines for a given knot vector using the Cox-de Boor 
recursion formula -- for knots $t_0$, $t_1$,...,$t_{i+k}$ the basis components are given by:
\begin{equation}
    B_{i,1}(x | \mathbf{t}) = 
    \begin{cases}
        1, & \text{if } t_i \leq x < t_{i+1} \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}
\begin{equation}
    B_{i,k+1}(x | \mathbf{t}) = \omega_{i,k}(x)B_{i,k}(x) + \big[1-\omega_{i+1,k}(x)\big]B_{i+1,k}(x)
\end{equation}
\noindent Where $\omega_{i,k}$ above is defined as:
\begin{equation}
    \omega_{i,k}(x | \mathbf{t}) =
    \begin{cases}
        \frac{x-t_i}{t_{i+k}-t_i}, & t_{i+k} \neq t_i \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}

A similar basis, the M-Spline basis, has desirable properties when looking to model a probability density function. Namely, 
$M_i \geq 0$ within $(t_i, t_i+k)$, zero elsewhere, and normalized to unity, $\int M_i(x)dx = 1$. The M-Spline basis only differs 
from B-Splines by a normalization factor, namely:
\begin{equation}\label{eq:MB_SplineRelation}
    M_{i,k} = \frac{k}{t_{i+k} - t_i} B_{i,k}
\end{equation}
\noindent We use this definition of the M-Spline basis for the models presented and applied in the later sections of this paper.

\begin{figure*}[ht!]
    \script{spline_basis_plot.py}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/spline_basis_plot.pdf}
        \caption{Plot showing proper (left) and improper (right) MSpline bases with 20 degrees of freedom and equal weights for each component. The 
            difference penalty as described in Section II hinges on the fact that basis coeficients that are close produce smoother or flatter 
            curves. This means one does not want to use an improper basis (i.e. right) with a difference penalty as the resulting curve is not flat 
            with each component having equal weights.}
        \label{fig:spline_basis}
    \end{centering}
\end{figure*}

Figure shows examples of two different M-Spline bases with 10 degrees of freedom and knots spaced linearly from 0 to 1. Notice 
that when we introduced knots and in equation \ref{eq:MB_SplineRelation}, you can see there needs to be knots up to $t_{i+k}$ while $i$ 
indexes the desired degrees of freedom in your basis. This means in practice one must define $n+k$ knots. However, you have control 
in the boundary conditions by how many of the $n+k$ knots you place at the lower and upper bounds. By placing $k$ knots at each boundary 
(leaving $n-k$ interior knots), the first and last basis components are maximized at the boundaries. Alternatively one may want to 
a-priori enforce vanishing boundary conditions, for example, if they knew for a fact that the function they are modeling has 
$p(x_\mathrm{bound}) = 0$.  This can be done by placing $k-1$ knots at the boundaries you wish the spline to vanish at. See figure for 
an example of a cubic M-Spline (k=3) basis with 10 degrees of freedom with $k=3$ knots at each boundary (left) versus enforcing vanishing 
boundary conditions by having $k-1=2$ knots at each boundary (right).

\subsection{Parameterized Perturbations} \label{sec:perturbation_model}
\subsection{Generic Mass Model} \label{sec:nonparametric_model}
\section{Results} \label{sec:results}

\begin{figure*}[ht!]
    \script{mass_distribution_plot.py}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/mass_distribution_plot.pdf}
        \caption{Plot showing the primary mass (left) and mass ratio (right) distributions inferred with the 
        \textsc{MSpline} model (black) with 40 knots for $m_1$, and 15 knots for $q$. We show the results for the \textsc{PowerlawPeak} (blue) and 
        \textsc{PowerlawSpline} (orange) models fromthe LVK's GWTC-3 population analyses. Solid lines show the median of the posterior while the shaded
        bands show the 90\% credible intervals.}
        \label{fig:mass_distribution}
    \end{centering}
\end{figure*}

\begin{figure*}[ht!]
    \script{component_spin_distribution_plot.py}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/component_spin_distribution_plot.pdf}
        \caption{Plot showing the inferred component spin magnitude (top) and spin tilt (bottom) distributions inferred with the 9-knot \textsc{MSpline} models. 
        There are two independent \textsc{MSpline} (black) models for each primary/secondary spin magnitude and spin tilt. We compare by showing the results for the
        \textsc{Default} (iid) spin model used with the \textsc{PowerlawPeak} mass model (blue) and the \textsc{PowerlawSpline} mass model (orange).}
        \label{fig:component_spin_distribution}
    \end{centering}
\end{figure*}

\subsection{Mass Spectrum Inference with GWTC-3} \label{sec:GWTC3_inferece}
\section{Conclusions}\label{sec:conclusion}
\section{Acknowledgements}\label{sec:acknowledments}
This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center 
(\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. 
This work benefited from access to the University of Oregon high performance computer, Talapas.  This material is based upon work supported 
in part by the National Science Foundation under Grant PHY-1807046 and work supported by NSF’s LIGO Laboratory which is a major facility 
fully funded by the National Science Foundation.
\software{
\textsc{Astropy}~\citep{2018AJ....156..123A},
\textsc{NumPy}~\citep{harris2020array},
\textsc{Matplotlib}~\citep{Hunter:2007},
\textsc{bilby}~\citep{Ashton_2019},
\textsc{GWPopulation}~\citep{Talbot_2019}
}
\bibliography{bib}{}
\bibliographystyle{aasjournal}

\appendix
\section{Hierarchical Bayesian Inference} \label{sec:hierarchical_inference}

% This appendix goes over in detail the hierarchical bayeisan inference framework
We use hierarchical Bayesian inference to simultaneously infer the population distributions of the primary masses ($m_1$), mass ratios 
($q$) and the redshifts ($z$) of observed BBHs. For a set of hyper-parameters, $\Lambda$, and local ($z=0$) merger rate density (units of 
mergers per co-moving volume per time), $\mathcal{R}_0$, we write the overall number density of BBH mergers in the universe as:

\begin{equation} \label{number_density}
     \frac{dN(m_1, q, z | \mathcal{R}_0, \Lambda)}{dm_1dqdz} = \frac{dV_c}{dz}\bigg(\frac{T_\mathrm{obs}}{1+z}\bigg) \frac{d\mathcal{R}(m_1, q, z | \mathcal{R}_0, \Lambda)}{dm_1dq} = \mathcal{R}_0 p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda)
\end{equation}

\noindent
Above, $dV_c$ is the co-moving volume element \citep{hogg_cosmo} and $T_\mathrm{obs}$, the observing time period that produced the 
catalog with the related factor of $1+z$ converting this detector-frame time to source-frame. We assume a LambdaCDM cosmology using 
the cosmological parameters from \citet{Planck2015}. The redshift evolution of the merger rate follows 
$p(z|\lambda) \propto \frac{dV_c}{dz}\frac{1}{1+z}(1+z)^\lambda$. Integrating equation \ref{number_density} across all masses, 
and up to some redshift, $z_m$, returns the total number of BBH mergers in the universe out to that redshift. Let $\{d_i\}$ represent a 
set of data from $N_\mathrm{obs}$ observed gravitational waves associated with BBH mergers. We model the merger rate as an inhomogenous 
point process and when imposing a log-uniform prior on the merger rate, we can marginalize over the merger rate to get the posterior 
distribution of our hyper-parameters, $\Lambda$ \citep{Mandel_2019, Vitale_2021}.

\begin{equation} \label{posterior}
    p\left(\Lambda | \{d_i\} \right) \propto \frac{p(\Lambda)}{\xi(\Lambda)^{N_\mathrm{obs}}} \prod_{i=1}^{N_\mathrm{obs}} \Bigg[ \int \mathcal{L}\left(d_i | m_1^i, q^i, z^i \right) p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda) dm_1 dq dz \Bigg],
\end{equation}

\noindent
where, $\mathcal{L}(d_i|m_1, q, z)$, is the single-event likelihood function from each events original analysis, and $\xi(\Lambda)$ 
is the detection efficiency given a population distribution described by $\Lambda$. The procedure for calculating $\xi(\Lambda)$ is 
described in more detail below. The LVC reports out posterior samples for each observed event, with which we can use importance sampling 
to estimate the integrals above in equation \ref{posterior}. We replace the integrals with ensemble averages over $K_i$ posterior samples 
associated with each event in the catalog:

\begin{equation}
 p\left(\Lambda | \{d_i\}\right) \propto p(\Lambda) \prod_{i=1}^{N_\mathrm{obs}} \bigg[ \frac{1}{K_i} \sum_{j=1}^{K_i} \frac{p(m_1^{i,j}|\Lambda)p(q^{i,j}|\Lambda)p(z^{i,j}|\Lambda)}{\pi(m_1^{i,j}, q^{i,j}, z^{i,j})} \bigg]
\end{equation}

\noindent 
Here $j$ indexes the $K_i$ posterior samples from each event and $\pi(m_1, q, z)$ is the default prior used by parameter estimations that 
produced the posterior samples for each event. In the analyses of GWTC-2 the default prior used is uniform in detector frame masses and 
Euclidean volume. The corresponding prior evaluated in source frame masses and redshift is 
$\pi(m_1, q, z) \propto m_1(1+z)^2 D_L^2(z) \frac{dD_L}{dz}$, where $D_L$ is the luminosity distance.

To carefully incorporate selection effects to our model we need to quantify the detection efficiency, $\xi(\Lambda)$, of the search 
pipelines that were used to create GWTC-2, at a given population distribution described by $\Lambda$.

\begin{equation} \label{det_eff}
     \xi(\Lambda) = \int dm_1 dq dz P_\mathrm{det}(m_1, q, z)p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda)
\end{equation}

\noindent The above integral is not tractable since there is no analytic prescription for $P_\mathrm{det}(m_1,q,z)$, the detection 
probability of an individual event. To estimate this integral we use a software injection campaign where GWs from a fixed, broad 
population of sources are simulated, put into real detector data, and then run through the same search pipelines that were used to 
produce the catalog we are analyzing \footnote{For O3a we used the injection sets used by \citet{o3a_pop}, which can be found at 
https://dcc.ligo.org/LIGO-P2000217/public. For O1/O2 we used the mock injection sets used by \citet{o1o2_pop} which can be found at 
https://dcc.ligo.org/LIGO-P2000434/public}. With these search results in hand, we use importance sampling to evaluate the integral 
in equation \ref{det_eff}:

\begin{equation} \label{xi}
     \xi(\Lambda) = \frac{1}{N_\mathrm{inj}} \sum_{i=1}^{N_\mathrm{found}} \frac{p(m_1^i | \Lambda) p(q^i | m_1, \Lambda) p(z^i | \Lambda)}{p_\mathrm{inj}(m_1^i, q^i, z^i)}
\end{equation}

\noindent
Where the sum indexes only over the $N_\mathrm{found}$ injections that were successfully detected out of $N_\mathrm{inj}$ total injections, 
and $p_\mathrm{inj}(m_1, q, z)$ is the reference distribution from which the injections were drawn. Additionally, we follow the procedure 
outlined in \citet{Farr_2019} to marginalize the uncertainty in our estimate of $\xi(\Lambda)$ due to a finite number of simulated events. 
We make the assumption that repeated sampling of $\xi(\Lambda)$ will follow a normal distribution with 
$\xi(\Lambda) \sim \mathcal{N}(\mu(\Lambda), \sigma(\Lambda))$, where the mean, $\mu$, is the estimate from equation \ref{xi}, 
while the variance, $\sigma^2$, is defined as:

\begin{equation}
    \sigma^2(\Lambda) \equiv \frac{\mu^2(\Lambda)}{N_\mathrm{eff}} \simeq \frac{1}{N^2_\mathrm{inj}} \sum_{i=1}^{N_\mathrm{found}} \bigg[\frac{p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda)}{p_\mathrm{inj}(m_1, q, z)}\bigg]^2 - \frac{\mu^2(\Lambda)}{N_\mathrm{inj}}
\end{equation}

\noindent
Above we define $N_\mathrm{eff}$ as the effective number of independent draws contributing to the importance sampled estimate, 
in which we verify to be sufficiently high after re-weighting the injections to a given population (i.e. $N_\mathrm{eff} > 4N_\mathrm{obs}$). 
We write the hyper-posterior marginalized over the merger rate and uncertainty in estimation of $\xi(\Lambda)$, neglecting terms of 
$\mathcal{O}(N_\mathrm{eff}^{-2})$ or greater \citep{Farr_2019}, as:

\begin{equation}\label{importance-posterior}
    \log p\left(\Lambda | \{d_i\}\right) \propto \sum_{i=1}^{N_\mathrm{obs}} \log \bigg[ \frac{1}{K_i} \sum_{j=1}^{K_i} \frac{p(m_1^{i,j}|\Lambda)p(q^{i,j}|\Lambda)p(z^{i,j}|\Lambda)}{\pi(m_1^{i,j}, q^{i,j}, z^{i,j})} \bigg] -  \\
    N_\mathrm{obs} \log \mu + \frac{3N_\mathrm{obs} + N_\mathrm{obs}^2}{2N_\mathrm{eff}} + \mathcal{O}(N_\mathrm{eff}^{-2}).
\end{equation}

\noindent
We explicitly enumerate each of the models used in this work for $p(m_1|\Lambda)$, $p(q|m_1, \Lambda)$, and $p(z|\Lambda)$ along with 
their respective hyper-parameters and prior distributions in Table \ref{tab:model_params}. To calculate marginal likelihoods and draw 
samples of the hyper parameters from the hierarchical posterior distribution shown in equation \ref{importance-posterior}, we use the 
\textsc{Bilby} \citep{Ashton_2019, bilby_gwtc1} and \textsc{GWPopulation} \citep{Talbot_2019} Bayesian inference software libraries with 
the \textsc{Dynesty} dynamic nested sampling algorithm \citep{Speagle_2020}.

\section{Validation Studies} \label{sec:validation}

\end{document}

>>>>

@@

Observations of gravitational waves from compact binary coalescences (CBCs) is becoming a regular occurrence, 
recently producing catalogs of 90 such detections \bruce{cite GWTCs}. This growing catalog size is beginning to enable access to a more complete 
understanding of the population properties CBCs, which encodes the history of the formation and evolution of these extreme compact objects. 
In the coming fourth observing run of the LIGO-Virgo-KAGRA collaboration \bruce{cite detectors}, the catalog size is expected to increase by up to a factor of 5. 
Looking at population properties of compact objects lets us probe the complex physics common to all of our observations that governs the 
formation and evolution of merging compact objects. The history of these dense objects is imprinted on the properties of the population 
of CBCs, allowing us to learn about these not well understood ideas through careful evaluation of the enture distribution of observed GWs. 
The mass distribution of binary black holes has been used to measure cosmological parameters, constrain modified gravitational wave propagation, 
constrain a non-constant Planck mass, look for evidence of ultralight bosons through superradiance, put constraints on the uncertain nuclear reaction 
rates important for stellar evolution, look for primordial black holes, to constrain the neutron star equation of state, 
and to understand the fromation channels/environments of compact binary coalescences (CBCs). Through a better understanding of the mass, spin, 
and redshift distribution of compact binaries that will come with the increased catalog size, one can probe these dufferent physical phenoma with 
greater fidelity.

@@

### 2022-09-14 src/tex/ms.tex
update the git repo

\usepackage{showyourwork}
\usepackage[perpage]{footmisc}
\usepackage{colortbl}
\usepackage{lineno}
\usepackage{amsmath}
\usepackage{appendix}
\definecolor{Gray}{gray}{0.9}
\linenumbers

@@

\section{Introduction} \label{sec:intro}

Now that gravitational wave observations of compact binary coalescence is a regular occurrence, producing catalogs of nearly 
100 such detections, we now have access to a complete understanding of the population of compact binaries. In the coming fourth 
observing run of the LIGO-Virgo-KAGRA collaboration, the catalog size is expected to increase by a factor of 5. The extensive 
physics governing the formation and evolution of compact objects that end up merging is imprinted on the properties of the population 
of CBCs, allowing us to learn about these not well understood ideas through looking at the distribution of observed GWs. In particular 
the mass distribution of binary black holes has been shown to be very useful in measuring cosmological parameters, constraining modified 
gravitational wave propagation, constraining a running Planck mass, looking of evidence of ultralight bosons through superradiance, 
putting constraints on uncertain nuclear reaction rates important for stellar evolution, and even looking for primordial black holes. 
Through a better understanding of the mass, spin, and redshift distribution of compact binaries we can extend this be a useful tool 
in all the different physical applications mentioned above. In this paper we highlight how the use of basis-splines can be a useful 
model-agnostic modeling approach to the astrophysical distributions of compact binaries. We show that with a particular choice of basis, 
we can efficiently model the spin and/or mass distribution of compact binaries in GWTC-3 with spline regression in hierarchical 
Bayesian inference. We show our results are similar to those produced with the more simpler and motivated parametric models, 
while being capable of finding more subtle features should they appear in future catalogs. This method can be extended to higher-dimensional 
basis-splines to model correlations between different distributions which will be necessary to confidently disentangle differently 
competing formation channels should they exist.

\section{Methods} \label{sec:methods}
\subsection{Constructing A Basis} \label{sec:basis_splines}

A common non-parametric method in statistical applications is the use of B-Splines or basis splines. A spline function of order $k$, 
is a piece-wise polynomial of order $k-1$ polynomials stitched together from defined ``knot'' locations across the domain. 
They provide a useful and cheap way to interpolate generically smooth functions from a finite sampling of ``knot'' heights. 
B-Splines of order $k$ are a set of order $k-1$ polynomials that form a complete basis for any spline function of order $k$. 
Given an array of knot locations, $\mathbf{t}$ or knot vector, there exists a single unique linear combination of B-Splines for 
every possible spline function with knots, $\mathbf{t}$. One can construct the B-Splines for a given knot vector using the Cox-de Boor 
recursion formula -- for knots $t_0$, $t_1$,...,$t_{i+k}$ the basis components are given by:
\begin{equation}
    B_{i,1}(x | \mathbf{t}) = 
    \begin{cases}
        1, & \text{if } t_i \leq x < t_{i+1} \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}
\begin{equation}
    B_{i,k+1}(x | \mathbf{t}) = \omega_{i,k}(x)B_{i,k}(x) + \big[1-\omega_{i+1,k}(x)\big]B_{i+1,k}(x)
\end{equation}
\noindent Where $\omega_{i,k}$ above is defined as:
\begin{equation}
    \omega_{i,k}(x | \mathbf{t}) =
    \begin{cases}
        \frac{x-t_i}{t_{i+k}-t_i}, & t_{i+k} \neq t_i \\
        0, & \text{otherwise}
    \end{cases}
\end{equation}

A similar basis, the M-Spline basis, has desirable properties when looking to model a probability density function. Namely, 
$M_i \geq 0$ within $(t_i, t_i+k)$, zero elsewhere, and normalized to unity, $\int M_i(x)dx = 1$. The M-Spline basis only differs 
from B-Splines by a normalization factor, namely:
\begin{equation}\label{eq:MB_SplineRelation}
    M_{i,k} = \frac{k}{t_{i+k} - t_i} B_{i,k}
\end{equation}
\noindent We use this definition of the M-Spline basis for the models presented and applied in the later sections of this paper.

Figure shows examples of two different M-Spline bases with 10 degrees of freedom and knots spaced linearly from 0 to 1. Notice 
that when we introduced knots and in equation \ref{eq:MB_SplineRelation}, you can see there needs to be knots up to $t_{i+k}$ while $i$ 
indexes the desired degrees of freedom in your basis. This means in practice one must define $n+k$ knots. However, you have control 
in the boundary conditions by how many of the $n+k$ knots you place at the lower and upper bounds. By placing $k$ knots at each boundary 
(leaving $n-k$ interior knots), the first and last basis components are maximized at the boundaries. Alternatively one may want to 
a-priori enforce vanishing boundary conditions, for example, if they knew for a fact that the function they are modeling has 
$p(x_\mathrm{bound}) = 0$.  This can be done by placing $k-1$ knots at the boundaries you wish the spline to vanish at. See figure for 
an example of a cubic M-Spline (k=3) basis with 10 degrees of freedom with $k=3$ knots at each boundary (left) versus enforcing vanishing 
boundary conditions by having $k-1=2$ knots at each boundary (right).

\subsection{Parameterized Perturbations} \label{sec:perturbation_model}
\subsection{Generic Mass Model} \label{sec:nonparametric_model}
\section{Results} \label{sec:results}
\subsection{Mass Spectrum Inference with GWTC-3} \label{sec:GWTC3_inferece}
\section{Conclusions}\label{sec:conclusion}
\section{Acknowledgements}\label{sec:acknowledments}
This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center 
(\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. 
This work benefited from access to the University of Oregon high performance computer, Talapas.  This material is based upon work supported 
in part by the National Science Foundation under Grant PHY-1807046 and work supported by NSF’s LIGO Laboratory which is a major facility 
fully funded by the National Science Foundation.
\software{
\textsc{Astropy}~\citep{2018AJ....156..123A},
\textsc{NumPy}~\citep{harris2020array},
\textsc{Matplotlib}~\citep{Hunter:2007},
\textsc{bilby}~\citep{Ashton_2019},
\textsc{GWPopulation}~\citep{Talbot_2019}
}
\bibliography{references}{}
\bibliographystyle{aasjournal}

\appendix
\section{Hierarchical Bayesian Inference} \label{sec:hierarchical_inference}

% This appendix goes over in detail the hierarchical bayeisan inference framework
We use hierarchical Bayesian inference to simultaneously infer the population distributions of the primary masses ($m_1$), mass ratios 
($q$) and the redshifts ($z$) of observed BBHs. For a set of hyper-parameters, $\Lambda$, and local ($z=0$) merger rate density (units of 
mergers per co-moving volume per time), $\mathcal{R}_0$, we write the overall number density of BBH mergers in the universe as:

\begin{equation} \label{number_density}
     \frac{dN(m_1, q, z | \mathcal{R}_0, \Lambda)}{dm_1dqdz} = \frac{dV_c}{dz}\bigg(\frac{T_\mathrm{obs}}{1+z}\bigg) \frac{d\mathcal{R}(m_1, q, z | \mathcal{R}_0, \Lambda)}{dm_1dq} = \mathcal{R}_0 p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda)
\end{equation}

\noindent
Above, $dV_c$ is the co-moving volume element \citep{hogg_cosmo} and $T_\mathrm{obs}$, the observing time period that produced the 
catalog with the related factor of $1+z$ converting this detector-frame time to source-frame. We assume a LambdaCDM cosmology using 
the cosmological parameters from \citet{Planck2015}. The redshift evolution of the merger rate follows 
$p(z|\lambda) \propto \frac{dV_c}{dz}\frac{1}{1+z}(1+z)^\lambda$. Integrating equation \ref{number_density} across all masses, 
and up to some redshift, $z_m$, returns the total number of BBH mergers in the universe out to that redshift. Let $\{d_i\}$ represent a 
set of data from $N_\mathrm{obs}$ observed gravitational waves associated with BBH mergers. We model the merger rate as an inhomogenous 
point process and when imposing a log-uniform prior on the merger rate, we can marginalize over the merger rate to get the posterior 
distribution of our hyper-parameters, $\Lambda$ \citep{Mandel_2019, Vitale_2021}.

\begin{equation} \label{posterior}
    p\left(\Lambda | \{d_i\} \right) \propto \frac{p(\Lambda)}{\xi(\Lambda)^{N_\mathrm{obs}}} \prod_{i=1}^{N_\mathrm{obs}} \Bigg[ \int \mathcal{L}\left(d_i | m_1^i, q^i, z^i \right) p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda) dm_1 dq dz \Bigg],
\end{equation}

\noindent
where, $\mathcal{L}(d_i|m_1, q, z)$, is the single-event likelihood function from each events original analysis, and $\xi(\Lambda)$ 
is the detection efficiency given a population distribution described by $\Lambda$. The procedure for calculating $\xi(\Lambda)$ is 
described in more detail below. The LVC reports out posterior samples for each observed event, with which we can use importance sampling 
to estimate the integrals above in equation \ref{posterior}. We replace the integrals with ensemble averages over $K_i$ posterior samples 
associated with each event in the catalog:

\begin{equation}
 p\left(\Lambda | \{d_i\}\right) \propto p(\Lambda) \prod_{i=1}^{N_\mathrm{obs}} \bigg[ \frac{1}{K_i} \sum_{j=1}^{K_i} \frac{p(m_1^{i,j}|\Lambda)p(q^{i,j}|\Lambda)p(z^{i,j}|\Lambda)}{\pi(m_1^{i,j}, q^{i,j}, z^{i,j})} \bigg]
\end{equation}

\noindent 
Here $j$ indexes the $K_i$ posterior samples from each event and $\pi(m_1, q, z)$ is the default prior used by parameter estimations that 
produced the posterior samples for each event. In the analyses of GWTC-2 the default prior used is uniform in detector frame masses and 
Euclidean volume. The corresponding prior evaluated in source frame masses and redshift is 
$\pi(m_1, q, z) \propto m_1(1+z)^2 D_L^2(z) \frac{dD_L}{dz}$, where $D_L$ is the luminosity distance.

To carefully incorporate selection effects to our model we need to quantify the detection efficiency, $\xi(\Lambda)$, of the search 
pipelines that were used to create GWTC-2, at a given population distribution described by $\Lambda$.

\begin{equation} \label{det_eff}
     \xi(\Lambda) = \int dm_1 dq dz P_\mathrm{det}(m_1, q, z)p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda)
\end{equation}

\noindent The above integral is not tractable since there is no analytic prescription for $P_\mathrm{det}(m_1,q,z)$, the detection 
probability of an individual event. To estimate this integral we use a software injection campaign where GWs from a fixed, broad 
population of sources are simulated, put into real detector data, and then run through the same search pipelines that were used to 
produce the catalog we are analyzing \footnote{For O3a we used the injection sets used by \citet{o3a_pop}, which can be found at 
https://dcc.ligo.org/LIGO-P2000217/public. For O1/O2 we used the mock injection sets used by \citet{o1o2_pop} which can be found at 
https://dcc.ligo.org/LIGO-P2000434/public}. With these search results in hand, we use importance sampling to evaluate the integral 
in equation \ref{det_eff}:

\begin{equation} \label{xi}
     \xi(\Lambda) = \frac{1}{N_\mathrm{inj}} \sum_{i=1}^{N_\mathrm{found}} \frac{p(m_1^i | \Lambda) p(q^i | m_1, \Lambda) p(z^i | \Lambda)}{p_\mathrm{inj}(m_1^i, q^i, z^i)}
\end{equation}

\noindent
Where the sum indexes only over the $N_\mathrm{found}$ injections that were successfully detected out of $N_\mathrm{inj}$ total injections, 
and $p_\mathrm{inj}(m_1, q, z)$ is the reference distribution from which the injections were drawn. Additionally, we follow the procedure 
outlined in \citet{Farr_2019} to marginalize the uncertainty in our estimate of $\xi(\Lambda)$ due to a finite number of simulated events. 
We make the assumption that repeated sampling of $\xi(\Lambda)$ will follow a normal distribution with 
$\xi(\Lambda) \sim \mathcal{N}(\mu(\Lambda), \sigma(\Lambda))$, where the mean, $\mu$, is the estimate from equation \ref{xi}, 
while the variance, $\sigma^2$, is defined as:

\begin{equation}
    \sigma^2(\Lambda) \equiv \frac{\mu^2(\Lambda)}{N_\mathrm{eff}} \simeq \frac{1}{N^2_\mathrm{inj}} \sum_{i=1}^{N_\mathrm{found}} \bigg[\frac{p(m_1 | \Lambda) p(q | m_1, \Lambda) p(z | \Lambda)}{p_\mathrm{inj}(m_1, q, z)}\bigg]^2 - \frac{\mu^2(\Lambda)}{N_\mathrm{inj}}
\end{equation}

\noindent
Above we define $N_\mathrm{eff}$ as the effective number of independent draws contributing to the importance sampled estimate, 
in which we verify to be sufficiently high after re-weighting the injections to a given population (i.e. $N_\mathrm{eff} > 4N_\mathrm{obs}$). 
We write the hyper-posterior marginalized over the merger rate and uncertainty in estimation of $\xi(\Lambda)$, neglecting terms of 
$\mathcal{O}(N_\mathrm{eff}^{-2})$ or greater \citep{Farr_2019}, as:

\begin{equation}\label{importance-posterior}
    \log p\left(\Lambda | \{d_i\}\right) \propto \sum_{i=1}^{N_\mathrm{obs}} \log \bigg[ \frac{1}{K_i} \sum_{j=1}^{K_i} \frac{p(m_1^{i,j}|\Lambda)p(q^{i,j}|\Lambda)p(z^{i,j}|\Lambda)}{\pi(m_1^{i,j}, q^{i,j}, z^{i,j})} \bigg] -  \\
    N_\mathrm{obs} \log \mu + \frac{3N_\mathrm{obs} + N_\mathrm{obs}^2}{2N_\mathrm{eff}} + \mathcal{O}(N_\mathrm{eff}^{-2}).
\end{equation}

\noindent
We explicitly enumerate each of the models used in this work for $p(m_1|\Lambda)$, $p(q|m_1, \Lambda)$, and $p(z|\Lambda)$ along with 
their respective hyper-parameters and prior distributions in Table \ref{tab:model_params}. To calculate marginal likelihoods and draw 
samples of the hyper parameters from the hierarchical posterior distribution shown in equation \ref{importance-posterior}, we use the 
\textsc{Bilby} \citep{Ashton_2019, bilby_gwtc1} and \textsc{GWPopulation} \citep{Talbot_2019} Bayesian inference software libraries with 
the \textsc{Dynesty} dynamic nested sampling algorithm \citep{Speagle_2020}.

\section{Validation Studies} \label{sec:validation}

>>>>

\usepackage{showyourwork}
\usepackage[perpage]{footmisc}
\usepackage{colortbl}
\usepackage{lineno}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{appendix}
\definecolor{Gray}{gray}{0.9}
\linenumbers

@@

\input{intro.tex}
\input{methods.tex}
\input{results.tex}
\input{conclusion.tex}

\section{Acknowledgements}\label{sec:acknowledments}
This research has made use of data, software and/or web tools obtained from the Gravitational Wave Open Science Center 
(\url{https://www.gw-openscience.org/}), a service of LIGO Laboratory, the LIGO Scientific Collaboration and the Virgo Collaboration. 
This work benefited from access to the University of Oregon high performance computer, Talapas.  This material is based upon work supported 
in part by the National Science Foundation under Grant PHY-1807046 and work supported by NSF’s LIGO Laboratory which is a major facility 
fully funded by the National Science Foundation.
\software{
\textsc{Astropy}~\citep{2018AJ....156..123A},
\textsc{NumPy}~\citep{harris2020array},
\textsc{Matplotlib}~\citep{Hunter:2007},
\textsc{bilby}~\citep{Ashton_2019},
\textsc{GWPopulation}~\citep{Talbot_2019},
\textsc{jax},
\textsc{numpyro},
\textsc{arviz},
\textsc{seaborn},
\textsc{corner} \bruce{CITE ALL ABOVE}
}
\bibliography{bib}{}
\bibliographystyle{aasjournal}

\input{appendicies.tex}
